Hierarchical Bayesian regression modelling predicts biochemical rection energies and facilitates quantitative modelling of cell metabolism

### 2023-10-27 paper/main.tex
Update paper

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

@@

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Hierarchical Bayesian regression modelling predicts biochemical rection energies and facilitates quantitative modelling of cell metabolism},
  pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Hierarchical Bayesian regression modelling predicts biochemical
rection energies and facilitates quantitative modelling of cell
metabolism}
\author{Teddy Groves \and Nicholas Luke Cowie \and Lars Keld Nielsen}
\date{}

\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

A kinetic model of cellular metabolism aims to express what is known
about a cellular process in the form of an \emph{in silico}
representation of the underlying network of chemical reactions. Kinetic
models can be used to improve production of target molecules, determine
regulatory networks (Christodoulou et al. 2018) and identify potential
drug targets (DeBerardinis and Chandel 2016; Liberti et al. 2017).
However, the use of kinetic models in practice is hindered by their
dependence on noisy and uncertain information sources. Quantitative in
vivo measurements of chemical abundances, and in vitro measurements
relating to kinetic parameters, both contain vital information but are
notoriously inaccurate (REFERENCES). Practically useful kinetic
modelling therefore requires a principled statistical approach that
encompasses multiple possible model parameterisations.

Bayesian statistical inference can combine the structural information
implicit in kinetic models with knowledge about metabolic parameters and
information from omics measurements (P. A. Saa and Nielsen 2016;
Gopalakrishnan, Dash, and Maranas 2020). However, kinetic models pose
serious computational challenges for Bayesian inference (Gutenkunst et
al. 2007; A. Raue et al. 2010).

The scope of a kinetic model is defined by a stoichiometric matrix,
\(S\), in which rows represent metabolites, columns represent reactions,
and matrix elements \(s_{ij}\) represent the stoichiometric coefficient
of metabolite \(i\) in reaction \(j\). The change in metabolite
concentrations is

@@

Another advantage is that Bayesian inference problems are well-posed
even when not all parameters are strongly identified. Sloppy models in
which measurable quantities are sensitive to combinations of parameters
but not to individual marginal parameter values are ubiquitous in models
of biological systems {[}gutenkunst\_2007; White et al. (2016){]}. The
parameter correlation structure represents the set of potential models
that describe the observed data. As we demonstrate in our case study
below, capturing this correlation structure is difficult outside a
Bayesian context.

@@

The most popular algorithm for fitting Bayesian statistical models is
Markov Chain Monte Carlo (MCMC). Modern MCMC algorithms allow
exploration of high- dimensional posterior distributions, have robust
failure diagnostics (Vehtari et al. 2021) and can incorporate fast
numerical solvers, thereby making inference feasible for Bayesian
kinetic models. Nonetheless, the kinetic modelling literature reports an
aversion to MCMC, rooted mainly in concerns about sampling time and the
presumed difficulty of implementing the required statistical model
(Andreas Raue et al. 2013; P. A. Saa and Nielsen 2016). We are only
aware of two recent attempts to implement a Bayesian kinetic modelling
approach using MCMC. Stapor et al. (2018) fitted detailed kinetic models
using relatively inefficient MCMC algorithms that do not scale well to
high dimensional parameter spaces limiting the scope of modelling.
Conversely, St. John et al. (2018) utilises an efficient sampling
algorithm but uses approximate kinetics, namely lin-log kinetics Visser
and Heijnen (2003), limiting the scope of interpreting parameters and
inferring cellular behaviour in experimental conditions outside the
reference dataset.

There have also been efforts to implement Bayesian inference for kinetic
models without the use of MCMC. Examples of alternative inference
methods include variational inference as in St. John et al. (2018),
rejection sampling and approximate Bayesian computation P. A. Saa and
Nielsen (2016) and Laplace approximation, in which the Fisher
information matrix is used to calculate a normal approximation around
the maximum a posteriori parameter configuration (Liebermeister and Noor
2021; Gopalakrishnan, Dash, and Maranas 2020; Stapor et al. 2018;
Andreas Raue et al. 2013). Non-MCMC-based Bayesian kinetic models have
limited utility because they lack reliable diagnostic tools for
verifying that their results approximate the target posterior
distribution. This is a problem because realistic kinetic models tend to
induce highly correlated, non-Gaussian, joint probability distributions
(Gutenkunst et al. 2007; Stapor et al. 2018).

@@

We constructed a kinetic model of the methionine cycle in Maud's format
using the description in Korendyaseva et al. (2008). The ordinary
differential equation system describing this model is shown in
Equation\,\eqref{eq-meth-ode}.

After specifying the qualitative aspects of the kinetic model, we chose
parameter values to use as ground truth by Monte Carlo sampling using a
previous model of the methionine cycle as a starting point (see P. Saa
and Nielsen (2015) for this model).

We used these parameters to simulate steady states in a range of
plausible experimental conditions, again using P. Saa and Nielsen (2015)
as a starting point. These steady states were then used to generate
simulated measurements using the measurement model described below.

@@

We conducted adaptive Hamiltonian Monte Carlo sampling for the full and
missing- data datasets, obtaining XXX post-warmup samples after XXX of
total computation time. For the Laplace approximation comparison, we
used Maud's Laplace mode. Full details as well as instructions for
reproducing our analysis can be found at XXX.

@@

Running standard diagnostic checks indicated that the samples we
generated were from the target posterior distribution. The improved
\(\hat{R}\) statistic (Vehtari et al. 2021) for every sampled variable
was within 1\% of 1, indicating appropriate mixing within and between
Markov chains. Additionally, the number of effective samples was high,
indicating that we generated enough posterior samples to support
inferences about the bulks of the distributions of the sampled
parameters. Furthermore, we observed no post-warmup divergent
transitions, indicating that the sampler was able to transform the
log-posterior distribution, avoiding any regions with excessive
curvature that might inhibit exploration via HMC.

@@

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared with
the corresponding marginal prior distributions, even though these
parameters' joint posterior distribution contained enough information to
make accurate out of sample predictions. In some cases, there were
two-dimensional correlations such as the one shown in the bottom left of
figure 3; in this case the marginal distribution of the two parameters
is roughly banana shaped. More commonly, however, two- dimensional pair
plots were insufficient to reveal the underlying correlation structure,
as seen in the bottom-right plot in figure 3. This does not mean that
the parameters were uncorrelated, but rather that the correlations
involve more than two parameters.

Overall, our results show that Maud can fit a realistic pathway-sized
dataset. This was achieved without fixing the marginal values of kinetic
parameters: the information required to make good predictions was
contained in the correlation structure of the joint posterior
distribution. This finding is consistent with previous analyses of
biological systems that found they are ``sloppy'', that is, sensitive to
parameter combinations rather than marginal parameter values, with
important combinations, scales and regions of sensitivity being
difficult to ascertain in advance Gutenkunst et al. (2007), Poirier
(1998).

@@

Maud is a command line application implementing Bayesian inference for a
wide range of realistic kinetic models. Maud is written in Python (Van
Rossum and Drake 2009), designed for use on Windows, Macos and Linux,
registered on the Python Package Index as
\texttt{maud-metabolic-models}, documented at \textless https://maud-
metabolic-models.readthedocs.io\textgreater{} and actively developed and
maintained at \url{https://github.com/biosustain/Maud/}.

@@

Maud inputs are structured directories, somewhat inspired by the PEtab
format Schmiester et al. (2021). A Maud input directory must contain a
toml (Preston-Werner, Tom and Gedam, Pradyun 2020) file called
\texttt{config.toml} which gives the input a name, configures how Maud
will be run and tells Maud where to find the other files, allowing these
to have custom names. It must also include a file containing a kinetic
model definition, a file specifying information about parameters and a
file with information experiments. The required structure of these files
is documented at \textless https://
maud-metabolic-models.readthedocs.io/en/latest/inputting.html\textgreater.
The input is validated against a Pydantic (Pydantic developers 2022)
data model which can be inspected at
\url{https://github.com/biosustain/Maud/tree/main/maud/data_model}.

We chose to implement a custom input format despite the existence of
standard formats in similar areas, including SBML (Keating et al. 2020)
and PEtab (Schmiester et al. 2021). This choice was partly motivated by
the need to ensure flexibility as Maud was developed, but there are also
features of SBML and PEtab that make them structurally unsuitable in
this context. Our requirements for an input format included that it be
mathematics-free, so that all mathematical details are encapsulated in
source code, and that it has a detailed, verifiable structure. These
requirements made toml more attractive than SBML: toml is easier for
humans to read and edit and can straightforwardly be validated using
tools like pydantic. Further, an SBML representation of our desired
input would not contain differential equations. It would therefore not
be interoperable with most SBML-targeting software, which typically
assumes that differential equations are available and does not know
about Maud's structure.

@@

Each of the terms on the right-hand side of (REFER TO EQUATION) is a
function of \(C\) and \(\theta\). This idea is taken from Noor et al.
(2013). The terms usefully gather physically meaningful and conceptually
distinct factors contributing to reaction fluxes. \(Enzyme\) captures
the effect of enzyme concentration, \(k_{cat}\) that of enzyme
efficiency, \(Reversibility\) quantifies thermodynamic effects,
\(Saturation\) the effect of enzyme availability and \(Allostery\) the
effect of post translational modifications.

We used the model of enzyme saturation from Liebermeister, Uhlendorf,
and Klipp (2010) and the generalised Monod-Wyman-Changeux model of
Allosteric regulation introduced in (Monod, Wyman, and Changeux 1965;
Changeux 2013; Popova and Sel'kov 1975, 1979) and used more recently in
Matos et al. (2022). To capture the effect on enzyme activity of coupled
phosphorylation and dephosphorylation processes we developed a new
mathematical model inspired by the generalised MWC model of allosteric
regulation. Full details of all mathematical aspects of Maud's kinetic
model can be found in supplementary material section (REFERENCE).

@@

Except for metabolites' standard condition Gibbs energy changes of
formation, Maud uses independent normal prior distributions for
parameters that can in principle be both negative and positive. For
parameters that are constrained to be positive, Maud uses independent
log-normal distributions. Formation energy parameters have a
multivariate normal prior distribution. Location, scale and covariance
parameters for all these prior distributions can be chosen freely by the
user.

@@

All measurements are specific to an experimental condition - that is, a
case where the true state of the network, including knockouts, boundary
conditions and state variables as well as kinetic and thermodynamic
parameters, can safely be assumed to be the same. Maud's statistical
model allows for arbitrarily many experimental conditions, and for any
measurable quantity to be measured any number of times in any condition.

@@

Flux measurements, representing the results of quantitative fluxomics
analyses, are modelled using a likelihood function from a standard
linear regression model, as shown in equation Equation\,8 (REFERENCE).
Flux measurements can be obtained from the results of isotope labelling
experiments using metabolic flux analysis, for example as described in
(Young 2014). When entering flux measurements, it is important only to
specify measurements for a network's free fluxes, as the values of some
steady state fluxes in a metabolic network are constrained by others,
with the result that dependent fluxes cannot typically be measured
separately. If measurements of multiple dependent fluxes are entered,
information will inappropriately be double counted.

@@

Finally, the reader may wonder why we chose a linear regression model
for reaction flux measurements even though this creates the potential
for erroneous double counting and requires non-trivial upstream
modelling, as intracellular fluxes typically cannot be measured
directly, but must be inferred from isotope labelling experiments using
metabolic flux analysis (see REFERENCE). Ideally Maud's measurement
model for fluxes would extend from fluxes to the results of potential
labelling experiments, thereby removing the need for upstream analysis
and avoiding any double counting. This option has not yet been pursued,
again for the sake of simplicity.

@@

Maud uses the Python library click (Click Developers 2022) to implement
a command line interface. The command line interface loads input files
as Python dictionaries, which are parsed using the Python library
\texttt{toml} (Pearson 2020) and then validated and converted into
structured \texttt{MaudInput} objects using Pydantic (Pydantic
developers 2022). Maud's statistical model is implemented in the
probabilistic programming language Stan (Carpenter et al. 2017) and
accessed using the interface cmdstanpy (Stan Development Team 2022). For
posterior sampling, Maud uses the \texttt{MaudInput} to create a Stan
input json file and obtain configuration information for cmdstanpy,
which it uses to (if necessary) generate a model executable file and
then trigger posterior sampling using adaptive Hamiltonian Monte Carlo.
When sampling is complete, Maud converts to the output into the standard
format InferenceData using the Python library arviz (Kumar et al. 2019)
and saves it as a json file, along with some information for debugging.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[journal=,manuscript=]{achemso}
\usepackage[version=3]{mhchem}
\newcommand*\mycommand[1]{\texttt{\emph{#1}}}

@@

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Hierarchical Bayesian regression modelling predicts biochemical rection energies and facilitates quantitative modelling of cell metabolism},
  pdfauthor={Teddy Groves; Nicholas Luke Cowie; Lars Keld Nielsen},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\author{Teddy Groves}
\affiliation{ DTU Biosustain,  }

\email{tedgro@biosustain.dtu.dk}
\author{Nicholas Luke Cowie}
\affiliation{ DTU Biosustain,  }

\author{Lars Keld Nielsen}
\affiliation{ DTU Biosustain,  }

\title[]{Hierarchical Bayesian regression modelling predicts biochemical
rection energies and facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, frame hidden, breakable, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

A kinetic model of cellular metabolism aims to express what is known
about a cellular process in the form of an \emph{in silico}
representation of the underlying network of chemical reactions. Kinetic
models can be used to improve production of target molecules, determine
regulatory networks \citep{christodoulou_reserve_2018} and identify
potential drug targets
\citep{deberardinis_fundamentals_2016, Liberti2017}. However, the use of
kinetic models in practice is hindered by their dependence on noisy and
uncertain information sources. Quantitative in vivo measurements of
chemical abundances, and in vitro measurements relating to kinetic
parameters, both contain vital information but are notoriously
inaccurate (REFERENCES). Practically useful kinetic modelling therefore
requires a principled statistical approach that encompasses multiple
possible model parameterisations.

Bayesian statistical inference can combine the structural information
implicit in kinetic models with knowledge about metabolic parameters and
information from omics measurements
\citep{saa_construction_2016, gopalakrishnan_k-fit_2020}. However,
kinetic models pose serious computational challenges for Bayesian
inference \citep{gutenkunst_2007, raue_identifiability_2010}.

The scope of a kinetic model is defined by a stoichiometric matrix,
\(S\), in which rows represent metabolites, columns represent reactions,
and matrix elements \(s_{ij}\) represent the stoichiometric coefficient
of metabolite \(i\) in reaction \(j\). The change in metabolite
concentrations is:

@@

Another advantage is that Bayesian inference problems are well-posed
even when not all parameters are strongly identified. Sloppy models in
which measurable quantities are sensitive to combinations of parameters
but not to individual marginal parameter values are ubiquitous in models
of biological systems \citep{gutenkunst_2007, white_2016}. The parameter
correlation structure represents the set of potential models that
describe the observed data. As we demonstrate in our case study below,
capturing this correlation structure is difficult outside a Bayesian
context.

@@

The most popular algorithm for fitting Bayesian statistical models is
Markov Chain Monte Carlo (MCMC). Modern MCMC algorithms allow
exploration of high- dimensional posterior distributions, have robust
failure diagnostics
\citep{vehtariRankNormalizationFoldingLocalization2021} and can
incorporate fast numerical solvers, thereby making inference feasible
for Bayesian kinetic models. Nonetheless, the kinetic modelling
literature reports an aversion to MCMC, rooted mainly in concerns about
sampling time and the presumed difficulty of implementing the required
statistical model \citep{raue_joining_2013, saa_construction_2016}. We
are only aware of two recent attempts to implement a Bayesian kinetic
modelling approach using MCMC. \citet{stapor_pesto_2018} fitted detailed
kinetic models using relatively inefficient MCMC algorithms that do not
scale well to high dimensional parameter spaces limiting the scope of
modelling. Conversely, \citet{st.johnBayesianInferenceMetabolic2018}
utilises an efficient sampling algorithm but uses approximate kinetics,
namely lin-log kinetics \citet{visser_dynamic_2003}, limiting the scope
of interpreting parameters and inferring cellular behaviour in
experimental conditions outside the reference dataset.

There have also been efforts to implement Bayesian inference for kinetic
models without the use of MCMC. Examples of alternative inference
methods include variational inference as in
\citet{st.johnBayesianInferenceMetabolic2018}, rejection sampling and
approximate Bayesian computation \citet{saa_construction_2016} and
Laplace approximation, in which the Fisher information matrix is used to
calculate a normal approximation around the maximum a posteriori
parameter configuration
\citep{liebermeister_model_2021, gopalakrishnan_k-fit_2020, stapor_pesto_2018, raue_joining_2013}.
Non-MCMC-based Bayesian kinetic models have limited utility because they
lack reliable diagnostic tools for verifying that their results
approximate the target posterior distribution. This is a problem because
realistic kinetic models tend to induce highly correlated, non-Gaussian,
joint probability distributions
\citep{gutenkunst_2007, stapor_pesto_2018}.

@@

We constructed a kinetic model of the methionine cycle in Maud's format
using the description in \citet{korendyaseva_allosteric_2008}. The
ordinary differential equation system describing this model is shown in
Equation\,\eqref{eq-meth-ode}.

After specifying the qualitative aspects of the kinetic model, we
selected parameter values to use as ground truth by Monte Carlo sampling
using a previous model of the methionine cycle as a starting point (see
\citet{saa_general_2015} for this model).

We used these parameters to simulate steady states in a range of
plausible experimental conditions, again using \citet{saa_general_2015}
as a starting point. These steady states were then used to generate
simulated measurements using the measurement model described below.

@@

We conducted adaptive Hamiltonian Monte Carlo sampling for the full and
missing- -data datasets, obtaining XXX post-warmup samples after XXX of
total computation time. For the Laplace approximation comparison, we
used Maud's Laplace mode. Full details as well as instructions for
reproducing our analysis can be found at XXX.

@@

Running standard diagnostic checks indicated that the samples we
generated were from the target posterior distribution. The improved
\(\hat{R}\) statistic
\citep{vehtariRankNormalizationFoldingLocalization2021} for every
sampled variable was within 1\% of 1, indicating appropriate mixing
within and between Markov chains. Additionally, the number of effective
samples was high, indicating that we generated enough posterior samples
to support inferences about the bulks of the distributions of the
sampled parameters. Furthermore, we observed no post warm-up divergent
transitions, indicating that the sampler was able to transform the
log-posterior distribution, avoiding any regions with excessive
curvature that might inhibit exploration via HMC.

@@

Analysis of the posterior distributions for kinetic parameters indicated
that these are highly correlated. The marginal posterior distributions
for most kinetic parameters did not shrink significantly compared with
the corresponding marginal prior distributions, even though these
parameters' joint posterior distribution contained enough information to
make accurate out of sample predictions. In some cases, there were
two-dimensional correlations such as the one shown in the bottom left of
figure 3; in this case the marginal distribution of the two parameters
is roughly banana shaped. More commonly, however, two-dimensional pair
plots were insufficient to reveal the underlying correlation structure,
as seen in the bottom-right plot in figure 3. This does not mean that
the parameters were uncorrelated, but rather that the correlations
involve more than two parameters.

Overall, our results show that Maud can fit a realistic pathway-sized
dataset. This was achieved without fixing the marginal values of kinetic
parameters: the information required to make good predictions was
contained in the correlation structure of the joint posterior
distribution. This finding is consistent with previous analyses of
biological systems that found they are ``sloppy'', that is, sensitive to
parameter combinations rather than marginal parameter values, with
important combinations, scales and regions of sensitivity being
difficult to ascertain in advance \citet{gutenkunst_2007},
\citet{poirier_revising_1998}.

@@

Maud is a command line application implementing Bayesian inference for a
wide range of realistic kinetic models. Maud is written in Python
\citep{vanrossumPythonReferenceManual2009}, designed for use on Windows,
macOS and Linux, registered on the Python Package Index as
\texttt{maud-metabolic-models}, documented at \textless https://maud-
metabolic-models.readthedocs.io\textgreater{} and actively developed and
maintained at \url{https://github.com/biosustain/Maud/}.

@@

Maud inputs are structured directories, somewhat inspired by the PEtab
format \citet{SchmiesterSch2021}. A Maud input directory must contain a
toml \citep{preston-wernertomandgedampradyunTOMLSpecification0rc2020}
file called \texttt{config.toml} which gives the input a name,
configures how Maud will be run and tells Maud where to find the other
files, allowing these to have custom names. It must also include a file
containing a kinetic model definition, a file specifying information
about parameters and a file with information experiments. The required
structure of these files is documented at \textless https://
maud-metabolic-models.readthedocs.io/en/latest/inputting.html\textgreater.
The input is validated against a Pydantic
\citep{pydanticdevelopersPydantic2022} data model which can be inspected
at \url{https://github.com/biosustain/Maud/tree/main/maud/data_model}.

We chose to implement a custom input format despite the existence of
standard formats in similar areas, including SBML
\citep{keatingSBMLLevelExtensible2020} and PEtab
\citep{SchmiesterSch2021}. This choice was partly motivated by the need
to ensure flexibility as Maud was developed, but there are also features
of SBML and PEtab that make them structurally unsuitable in this
context. Our requirements for an input format included that it be
mathematics-free, so that all mathematical details are encapsulated in
source code, and that it has a detailed, verifiable structure. These
requirements made toml more attractive than SBML: toml is easier for
humans to read and edit and can straightforwardly be validated using
tools like Pydantic. Further, an SBML representation of our desired
input would not contain differential equations. It would therefore not
be interoperable with most SBML targeting software, which typically
assumes that differential equations are available and does not know
about Maud's structure.

@@

Each of the terms on the right-hand side of (REFER TO EQUATION) is a
function of \(C\) and \(\theta\). This idea is taken from
\citet{noor_note_2013}. The terms usefully gather physically meaningful
and conceptually distinct factors contributing to reaction fluxes.
\(Enzyme\) captures the effect of enzyme concentration, \(k_{cat}\) that
of enzyme efficiency, \(Reversibility\) quantifies thermodynamic
effects, \(Saturation\) the effect of enzyme availability and
\(Allostery\) the effect of post translational modifications.

We used the model of enzyme saturation from
\citet{liebermeister_modular_2010} and the generalised
Monod-Wyman-Changeux model of Allosteric regulation introduced in
\citep{monod_nature_1965, changeux_2013, popova_generalization_1975, popova_description_1979}
and used more recently in \citet{matosGRASPComputationalPlatform2022}.
To capture the effect on enzyme activity of coupled phosphorylation and
dephosphorylation processes we developed a new mathematical model
inspired by the generalised MWC model of allosteric regulation. Full
details of all mathematical aspects of Maud's kinetic model can be found
in supplementary material section (REFERENCE).

@@

Except for metabolites' standard condition Gibbs energy changes of
formation, Maud uses independent normal prior distributions for
parameters that can in principle be both negative and positive. For
parameters that are constrained to be positive, Maud uses independent
log-normal distributions. Formation energy parameters have a
multivariate normal prior distribution. Location, scale and covariance
parameters for all these prior distributions can be selected freely by
the user.

@@

All measurements are specific to an experimental condition; that is, a
case where the true state of the network, including knockouts, boundary
conditions and state variables as well as kinetic and thermodynamic
parameters, can safely be assumed to be the same. Maud's statistical
model allows for arbitrarily many experimental conditions, and for any
measurable quantity to be measured any number of times in any condition.

@@

Flux measurements, representing the results of quantitative fluxomics
analyses, are modelled using a likelihood function from a standard
linear regression model, as shown in equation\,8 (REFERENCE). Flux
measurements can be obtained from the results of isotope labelling
experiments using metabolic flux analysis, for example as described in
(Young 2014). When entering flux measurements, it is important only to
specify measurements for a network's free fluxes, as the values of some
steady state fluxes in a metabolic network are constrained by others,
with the result that dependent fluxes cannot typically be measured
separately. If measurements of multiple dependent fluxes are entered,
information will inappropriately be double counted.

@@

Finally, the reader may wonder why Maud uses a linear regression model
for reaction flux measurements even though this creates the potential
for erroneous double counting and requires non-trivial upstream
modelling, as intracellular fluxes typically cannot be measured
directly, but must be inferred from isotope labelling experiments using
metabolic flux analysis (see REFERENCE). Ideally Maud's measurement
model for fluxes would extend from fluxes to the results of potential
labelling experiments, thereby removing the need for upstream analysis
and avoiding any double counting. This option has not yet been pursued,
again for the sake of simplicity.

@@

Maud uses the Python library click
\citep{clickdevelopersClickPythonComposable2022} to implement a command
line interface. The command line interface loads input files as Python
dictionaries, which are parsed using the Python library \texttt{toml}
\citep{pearsonTomlPythonLibrary2020} and then validated and converted
into structured \texttt{MaudInput} objects using Pydantic
\citep{pydanticdevelopersPydantic2022}. Maud's statistical model is
implemented in the probabilistic programming language Stan
\citep{carpenterStanProbabilisticProgramming2017} and accessed using the
interface cmdstanpy \citep{standevelopmentteamCmdStanPy2022}. For
posterior sampling, Maud uses the \texttt{MaudInput} to create a Stan
input json file and obtain configuration information for cmdstanpy,
which it uses to (if necessary) generate a model executable file and
then trigger posterior sampling using adaptive Hamiltonian Monte Carlo.
When sampling is complete, Maud converts to the output into the standard
format \texttt{InferenceData} using the Python library arviz
\citep{kumarArviZUnifiedLibrary2019} and saves it as a json file, along
with some information for debugging.
